\Title

### 2015-07-30 response_to_reviewers.txt

##############################################################################
# Reviewer #1 (Comments for Author):
##############################################################################

@@

>These are interesting results and the manuscript is clearly written. The second
>observation is not novel, but underappreciated, whereas this is the first
>concrete calculation of the average selective disadvantage of silent mutations
>in HIV taking draft into account. As such, it is necessary for this work to
>establish proper error estimates so that the community can make valid use of
>the results.  With this in mind, the manuscript can be strengthened in the
>following respects:

We thank the reviewer for the constructive comments, which have helped
us to improve the manuscript. We have resorted to a more sophisticated
sampling scheme for parameters in the computer simulations and now scan
a larger parameter space. All our conclusions remain valid, but are based
on more exhaustive sampling. We detail these improvements in our point to point
response below.

##############################################################################
>1) The modeling generation time of 1 day is odd.  It takes about a day for the
>virion to integrate and start transcription.  One usually takes 2 days as the
>mean generation time, but there is individual variation. See next point.

We agree that one generation in HIV takes closer to 2 days than 1
day. Most escape rates and selection coefficients, however, are more
readily measured in days and we have chosen to model HIV evolution with
generation of one day for simplicity. We have repeated some of the
simulations with a generation time of 2 days and find the same results.
In fact, minor changes in the generation have little influence on the
dynamics and only renormalize genetic drift. Since the main determinant
of genetic drift -- the (effective) population size -- is not known
accurately, we don't think the difference of choosing one or two days as
generation time is important. Our more exhaustive survey of the
parameter space includes a 5-fold variation of the population size. Our
results are very insensitive to the population size and hence also
insensitive to the generation time.

@@

##############################################################################
>2) Related to the above: the simulation does not address how sensitive the
>results are to varying the parameters (in particular, the generation time, the
>outcrossing rate, the mutation rate, and the assigned deleterious fitness of
>the "non-synonymous" changes).  In the absence of described theoretical
>analysis shaping our expectations, I strongly feel that this sensitivity
>analysis needs to be part of the results, and folded into the error bars.

We have greatly extended the parameter space that we survey and now test
different values for the mutation rate, the recombination rate, the
fitness effects of synonymous and non-synonymous mutations, and the
population size. To accomodate this high dimensional data, we have
changed figure 4. The figure now shows the joint distribution of the
deleterious effect and the fraction of synonymous mutations that are
deleterious that are compatible with observations as a contour plot. The
marginal distributions of the individual parameters consistent with the
data are shown in the supplement.

The distribution of compatible deleterious effects is again peaked around
0.002 and the fraction of "synonymous" mutations that can be neutral is
low, as reported in the earlier version of our manuscript.

@@

We agree with reviewer that HIV evolution is bursty, but we would like
to mention that actual number of cells that get infected every day is
around 10^7 and hence much larger than we are able to simulate. The
population size in our simulation can therefore be viewed as an
effective population size. We have investigated the sensitivity of our
results to changes in the population size and find that our conclusions
are insensitive to the population size as long as the product of
population size and mutation rate is not very much smaller than 1.

We have added a discussion of our choice of population size to the
methods section next the part where the other parameter choices are
discussed.

##############################################################################
>6) In the discussions, the rise of synonymous mutations is attributed to draft,
>but the fall to selective effects. In general, some of the fall might also be
>due to overlapping selective sweeps: the manuscript does not discuss this
>possibility.

Yes, much of the stochastic component in the dynamics of the frequencies
of synonymous mutations will be due to draft. However, the consistent
loss of variants, conditional on them being at high frequency earlier,
implies selection against these variants. Without a deleterious effect,
the mutations would fix with a probability equal to their frequency,
regardless of whether they are subject to drift of hitchhiking. We have
added a sentence to the discussion to make this point clear.

### 2015-07-30 supplement_body.tex
fixed supplementary marginals

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Time-dependent selection}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[h]
\begin{center}
\includegraphics[width=0.6\linewidth]{simulations_gradually_timeselec_deadagain}
\caption{
Time-dependent selection reduces fixation of nonsynonymous SNVs. The figure
compares the fixation probability in the time independent model (na\"ive) to
a model with time dependent selection that mimics  an evolving immune system.
It has been found that virus is typically neutralized by serum from a few months
earlier~\citep{richman_rapid_2003} but not by contemporary serum. We model this
evolving immune system by assuming that escaped variants lose their beneficial
effect with a rate proportional to the frequency of the escaped variant. 
Specifically, the selection effect of the escape mutations is
reset to its fitness cost of $-0.02$ with probability
\[ P_\text{recognized}(t) = c \cdot \nu(t), \] 
per generation, where $c$ is a constant coefficient shown in the legend that
encodes the overall efficiency of the host immune system. With increasing
probability of recognition, the fixation of frequent escape mutants is reduced,
while hitch-hiking of synonymous SNVs is not affected. The precise
shape of $\pfix(\nu)$ depends on the details of the $P_\text{recognized}(t)$, and 
we do not think that the high $\pfix(\nu)$ for $\nu<0.2$ is meaningful.
The other parameters for the shown simulations are
the following: deleterious effect $s_d = 10^{-3}$, average escape rate $\epsilon = 0.016$,
fraction of deleterious synonymous mutations $\alpha = 0.986$, rate of new epitopes
$k_A=0.0014$ per generation.
}
\label{fig:tds}
\end{center}
\end{figure}

\newpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Within-epitope competition}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[h]
\begin{center}
\includegraphics[width=0.6\linewidth]{simulations_gradually_epitopes}
\caption{
Competition between escape SNVs in the same epitope reduces fixation of
nonsynonymous SNVs. The figure compares the fixation probability of models
with one, three, or six mutually exclusive escape mutations within the
same epitope. Within epitope competition results in reduced fixation
probabilities of nonsynonymous changes, whereas the synonymous changes behave 
similarly in all cases. We assume that escape can happen at $n$ sites out of 3
consecutive codons and vary $n$.
The fitness landscape of each epitope includes negative epistatic terms, so that
the joint presence of more than one escape mutation is not any more beneficial
for the virus than a single mutation. Specifically, each site has two alleles,
$\pm 1$, where $-1$ is the ancestral one and $+1$ the derived one; the fitness
coefficient of a $k$-tuple of sites within the epitope is $f_k = (-1)^{k-1}
2^{1-n}\eta_\epsilon $, where $\eta_\epsilon$ is the escape rate of the epitope
drawn from an exponential distribution with mean $\epsilon$ and 
$n$ is the number of competing escapes in the epitope. 
In this evolutionary scenario, many escape SNVs start to sweep on different backgrounds within the viral population, but eventually
compete and only one of them fixes. The other parameters for the shown simulations are
the following: deleterious effect $s_d = 10^{-3}$, average escape rate $\epsilon = 0.016$,
fraction of deleterious synonymous mutations $\alpha = 0.986$, rate of new epitopes
$k_A=0.0014$ per generation.
}
\label{fig:wec}
\end{center}
\end{figure}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Time-dependent selection}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[h]
\begin{center}
\includegraphics[width=0.6\linewidth]{simulations_gradually_timeselec_deadagain}
\caption{
Time-dependent selection reduces fixation of nonsynonymous SNVs. The figure
compares the fixation probability in the time independent model (na\"ive) to
a model with time dependent selection that mimics  an evolving immune system.
It has been found that virus is typically neutralized by serum from a few months
earlier~\citep{richman_rapid_2003} but not by contemporary serum. We model this
evolving immune system by assuming that escaped variants lose their beneficial
effect with a rate proportional to the frequency of the escaped variant. 
Specifically, the selection effect of the escape mutations is
reset to its fitness cost of $-0.02$ with probability
\[ P_\text{recognized}(t) = c \cdot \nu(t), \] 
per generation, where $c$ is a constant coefficient shown in the legend that
encodes the overall efficiency of the host immune system. With increasing
probability of recognition, the fixation of frequent escape mutants is reduced,
while hitch-hiking of synonymous SNVs is not affected. The precise
shape of $\pfix(\nu)$ depends on the details of the $P_\text{recognized}(t)$, and 
we do not think that the high $\pfix(\nu)$ for $\nu<0.2$ is meaningful.
The other parameters for the shown simulations are
the following: deleterious effect $s_d = 0.002$, average escape rate $\epsilon
= 0.032$ per day, fraction of deleterious synonymous mutations $\alpha = 0.986$, rate of new epitopes
$k_A=0.0014$ per day.
}
\label{fig:tds}
\end{center}
\end{figure}

\newpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Within-epitope competition}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[h]
\begin{center}
\includegraphics[width=0.6\linewidth]{simulations_gradually_epitopes}
\caption{
Competition between escape SNVs in the same epitope reduces fixation of
nonsynonymous SNVs. The figure compares the fixation probability of models
with one, three, or six mutually exclusive escape mutations within the
same epitope. Within epitope competition results in reduced fixation
probabilities of nonsynonymous changes, whereas the synonymous changes behave 
similarly in all cases. We assume that escape can happen at $n$ sites out of 3
consecutive codons and vary $n$.
The fitness landscape of each epitope includes negative epistatic terms, so that
the joint presence of more than one escape mutation is not any more beneficial
for the virus than a single mutation. Specifically, each site has two alleles,
$\pm 1$, where $-1$ is the ancestral one and $+1$ the derived one; the fitness
coefficient of a $k$-tuple of sites within the epitope is $f_k = (-1)^{k-1}
2^{1-n}\eta_\epsilon $, where $\eta_\epsilon$ is the escape rate of the epitope
drawn from an exponential distribution with mean $\epsilon$ and 
$n$ is the number of competing escapes in the epitope. 
In this evolutionary scenario, many escape SNVs start to sweep on different backgrounds within the viral population, but eventually
compete and only one of them fixes. The other parameters for the shown simulations are
the following: deleterious effect $s_d = 0.002$, average escape rate
$\epsilon = 0.032$ per day, fraction of deleterious synonymous mutations $\alpha = 0.986$, rate of new epitopes
$k_A=0.0014$ per day.
}
\label{fig:wec}
\end{center}
\end{figure}

\newpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\section{Simulations with different parameters}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% \begin{figure}[h]
% \begin{center}
% \includegraphics[width=0.4\linewidth]{simulations_syn_gentime_2days.pdf}\\
% \includegraphics[width=0.24\linewidth]{simulations_syn_mu_1e-5.pdf}
% \includegraphics[width=0.24\linewidth]{simulations_syn_latin_suppl.pdf}
% \includegraphics[width=0.24\linewidth]{simulations_syn_mu_3e-5.pdf}
% \includegraphics[width=0.24\linewidth]{simulations_syn_mu_4e-5.pdf}\\
% \includegraphics[width=0.32\linewidth]{simulations_syn_coi_5e-3.pdf}
% \includegraphics[width=0.32\linewidth]{simulations_syn_latin_suppl2.pdf}
% \includegraphics[width=0.32\linewidth]{simulations_syn_coi_2e-2.pdf}
% \caption{
% Results of simulations with alternative parameters.
% \textbf{Panel A)} Like \figurename~4B in the main text, but simulations are performed using a
% generation time of two days. Two selection coefficient of synonymous sites
% doubles in units of 1/generations, hence keeps constants in units of 1/days.
% \textbf{Panels B, C, D, and E)} Effect of mutation rate on the simulation
% results. The mutation rates are respectively $1 \times 10^{-5}$ (B), $2 \times
% 10^{-5}$ (C), $3 \times 10^{-5}$ (D), and $4 \times
% 10^{-5}$ (E) per generation per site. Increasing the mutation rate tightens the
% requirement that almost all synonymous mutations must be deleterious. This can
% be rationalized as follows: a higher $\mu$ increases the neutral synonymous
% diversity (we have measured this directly in our simulations), hence more and
% more synonymous SNVs that reach high frequency are neutral and the
% effect of deleterious SNVs, although present, becomes too weak to be observed as
% a depression in $P_\text{fix}$. As a consequence, only simulations with almost
% no neutral sites still show a reduced $P_\text{fix}$ and are shown in the
% figures.
% \textbf{Panels F, G, H)} Effect of recombination rate on the simulation results.
% The recombination rates are respectively $0.5 \times 10^{-5}$ (F), $10^{-5}$
% (G), and $2 \times 10^{-5}$ (H) per generation per site. Recombination does not
% change the results qualitatively in terms of fixation probability. Simulations
% with high recombination rates, however, have an excessively high synonymous
% diversity compared to the data (dots at the bottom and right side of panel H).
% More stringent gating around the data solves this issue.
% }
% \label{fig:sims_altparameters}
% \end{center}
% \end{figure}

\newpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Marginal distributions in simulations}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[h]
\begin{center}
\includegraphics[width=0.99\linewidth]{simulations_marginals_stratified.pdf}
\caption{
Marginal distributions of simulation parameters that reproduce the
synonymous fixation probability and diversity seen in the data. Each
panel contains the prior distribution of parameters (prior), the distribution of
parameters after imposing the lower or upper bounds on synonymous
diversity (syn diversity $\gtrless$), the bounds on the fixation
probability (Area of $P_{fix}$), and the all bounds
simultaneously (gated). The distribution of population sizes, the recombination rates, the escape
rate of new epitopes, and the mutation rate remain approximately
flat. They only two directions parameters whose distribution is strongly
constrained by the data is the cost of synonymous mutations and the
fraction of synonymous mutations that are neutral. We note that the
distribution of compatible parameters is most likely a complicated
subspace of the high dimensional parameter space and interpretation of
these marginal projections is difficult. 
}
\label{fig:sims_altparameters}
\end{center}
\end{figure}

% \newpage
% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% \section{Divergence and diversity over time in simulations}
% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% \begin{figure}[h]
% \begin{center}
% \includegraphics[width=0.6\linewidth]{simulations_diversity_divergence_example.pdf}
% \caption{
% Diversity and divergence over time in an example simulation with realistic
% parameters. Divergence from the founder genotype accumulates linearly  in time. On the
% contrary, because of the frequent selective sweeps of escape mutations, diversity saturates
% relatively quickly. The population dynamics after that point is in a steady
% state.
% }
% \label{fig:sims_altparameters}
% \end{center}
% \end{figure}
